---
title: "Exploratory Data Analysis"
author: "Kailen Shantz"
output:
  html_notebook: default
  pdf_document: default
---

Last updated: `r {Sys.Date()}`


In this notebook, I'll be exploring the data that has been scraped from the web 
and cleaned to get a sense of what features might do well at predicting a 
season's winner, how the data should be modeled, and what types of feature 
transforms might be useful.

```{r "setup-workspace", message=F, echo=T,  warning=F}
# load libraries
library(here)
library(tidyverse)
library(forcats)
library(ggbeeswarm)

# load data
rpdr <- readr::read_delim(here("data/processed/rpdr_df.txt"), delim = "\t")
```

To start, we'll take a look at the structure of the data and some summary statistics.

```{r}
str(rpdr)
```
Unsurprisingly, `season` has been read in as a numeric variable. This shouldn't 
matter, but just to be safe let's change this to a factor. Note that I'm
choosing a factor instead of a character vector so that the order of seasons 
will be preserved.

```{r}
rpdr$season <- forcats::as_factor(as.character(rpdr$season))
```


```{r}
summary(rpdr)
```
Overall, these look like reasonable values, suggesting that nothing has gone
wrong in the data processing steps. It is notable that the max values for being
in the top, in the bottom and winning the mini challenges are rather high 
compared to the means. So let's take a closer look at these.

```{r}
rpdr$contestant[(rpdr$n_in_top == 8)]
rpdr$contestant[(rpdr$n_in_bottom == 7)]
rpdr$contestant[(rpdr$mini_challenge_wins == 5)]
```
Looking at the Wikipedia pages for each contestant's season, it appears that
these data are, in fact correct.


# Exploring the raw data

Next, let's get a sense of how the data for each performance metric is
distributed, and where the winners tend to fall.

```{r}
ggplot(rpdr, aes(x = n_lipsync, fill = was_winner)) +
  geom_histogram(binwidth = 1) +
  theme_bw()
```

Unsurprisingly, the season winners tended to have fewer lipsyncs, and half of
the time, the winner never had  to lipsync. No winner has ever had to lipsync
more than once. So this is likely a helpful  predictor. But it's also far from
sufficient, given how many non-winners either never had to lipsync, or only
had one lipsync. 

This also doesn't capture the fact that some people who had to
lipsync once also went home that very episode. A more useful way to look at this
data might be to examine times in the bottom vs number of lipsyncs. (later)

```{r}
ggplot(rpdr, aes(x = n_in_bottom, fill = was_winner)) +
  geom_histogram(binwidth = 1) +
  theme_bw()
```

Again, no surprises here. The winners tended to be in the bottom few times. But
again, we can also see that this is far from a perfect predictor of the winner.

```{r}
ggplot(rpdr, aes(x = n_safe, fill = was_winner)) +
  geom_histogram(binwidth = 1) +
  theme_bw()
```

This looks like a less informative predictor. The distribution appears to be
relatively uniform. The winners are spread across all possible values for the 
number of time safe except 0. So  minimnally, we can say that all winners have 
been safe at least once.

```{r}
ggplot(rpdr, aes(x = n_in_top, fill = was_winner)) +
  geom_histogram(binwidth = 1) +
  theme_bw() +
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8))
```

Nothing here should be particularly surprising. Season winners tended to be in
the top a fair number of times, and have thus far never been in the top fewer
than 3 times.


```{r}
ggplot(rpdr, aes(x = n_wins, fill = was_winner)) +
  geom_histogram(binwidth = 1) +
  theme_bw()
```

Just like we saw for times in the top, season winners fall on the right tail
of this distribution, always having multiple wins.

It's worth considering that using count data might be a little misleading. On
the one hand, the winners tend to cluster together in the histograms, but we
did observe some separation for the winners when it came to how often they
were in the top. One potential problem is that there are a different number of
episodes and different numbers of competitors across seasons. And using pure
count data doesn't take that variation into account. So for starters, let's take
a look at how these distributions vary across seasons.

```{r, fig.height=7, fig.width=2, fig.asp=0.9}
ggplot(rpdr, aes(x = n_lipsync, fill = was_winner)) +
  geom_histogram(binwidth = 1) +
  facet_grid(season~.) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 5))
```

These distributions actually look quite comparable across seasons. We can see,
now, that the winner was not always the person with the fewest lipsyncs. But,
again, this is a limitation of using count data. It doesn't capture facts like 
how a contestant can lipsync once and then be sent home that same episode.

```{r, fig.height=7, fig.width=2, fig.asp=0.9}
ggplot(rpdr, aes(x = n_in_bottom, fill = was_winner)) +
  geom_histogram(binwidth = 1) +
  facet_grid(season~.) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 5))
```

For the most part, the data are similar across seasons. Times in the bottom
tend to be between 1 and 4. Seson 1 has a tighter spread, likely because it had
fewer contestants. Seasons 6 and 7 stand out a bit with more apparent right
skews in the data.

```{r, fig.height=7, fig.width=2, fig.asp=0.9}
ggplot(rpdr, aes(x = n_safe, fill = was_winner)) +
  geom_histogram(binwidth = 1) +
  facet_grid(season~.) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 5))
```

For the most part, these data look consistent with what we saw in the aggregated
data: there is a relitively uniform distribution in each seasons. One thing that
does stand out is that as we go from earlier to later seasons, the number of
times the winner is safe has increased.

```{r, fig.height=7, fig.width=2, fig.asp=0.9}
ggplot(rpdr, aes(x = n_in_top, fill = was_winner)) +
  geom_histogram(binwidth = 1) +
  facet_grid(season~.) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 5)) + 
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8))
```

Here we can see that while there is variation across seasons, for the most part,
the winner tends to be consistently in the top.

```{r, fig.height=7, fig.width=2, fig.asp=0.9}
ggplot(rpdr, aes(x = n_wins, fill = was_winner)) +
  geom_histogram(binwidth = 1) +
  facet_grid(season~.) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 5)) + 
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8))
```

These plot reveals more to us than the aggregate plot. Here we can very cleary
see that, with the exception of season 9, the winner consistently has either
the most wins, or is tied with one or two other contestants for the most wins.

Now that we've got a sense of what the data look like, let's take a deeper look
at each season to see how the competitors stack up against one another. To do
this, I'll be plotting stacked barcharts for how often each participant in a
season lipsynced, was in the bottom, was safe, was in the top and won.

Note that times in the bottom includes the times that a participant had to 
lispync. So, for example the first person to get sent home would be in the 
bottom the same number of times that they lipsynced.

Similarly, times in the top includes wins. So if a participant was in the top
the same number of times that they won, this means that they won every time they
were in the top.

```{r}
rpdr %>% gather(key = "performance", value = "count", n_lipsync:n_wins) %>%
  filter(season == 1) %>%
  ggplot(aes(x = reorder(contestant, n_appearances), y = count, fill = fct_relevel(performance, "n_lipsync", "n_in_bottom", "n_safe", "n_in_top", "n_wins"))) +
  geom_col(alpha = 0.7) + 
  coord_flip() + 
  labs(y = "Count", fill  = "Performance", x = "Contestants", title = "Season 1") +
  theme_bw() + scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18), expand = c(0,0))
```

```{r}
rpdr %>% gather(key = "performance", value = "count", n_lipsync:n_wins) %>%
  filter(season == 2) %>%
  ggplot(aes(x = reorder(contestant, n_appearances), y = count, fill = fct_relevel(performance, "n_lipsync", "n_in_bottom", "n_safe", "n_in_top", "n_wins"))) +
  geom_col(alpha = 0.7) + 
  coord_flip() + 
  labs(y = "Count", fill  = "Performance", x = "Contestants", title = "Season  2") +
  theme_bw() + scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18), expand = c(0,0))
```

```{r}
rpdr %>% gather(key = "performance", value = "count", n_lipsync:n_wins) %>%
  filter(season == 3) %>%
  ggplot(aes(x = reorder(contestant, n_appearances), y = count, fill = fct_relevel(performance, "n_lipsync", "n_in_bottom", "n_safe", "n_in_top", "n_wins"))) +
  geom_col(alpha = 0.7) + 
  coord_flip() + 
  labs(y = "Count", fill  = "Performance", x = "Contestants", title = "Season  3") +
  theme_bw() + scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18), expand = c(0,0))
```

This graph is great because it highlights a point where any model fit to these
data will fail. Manila's and Raja's performance was identical, so the best we
can accomplish with our current measures is a model that predicts they will have
an equally likely chance of winning season 3. In other words, the performance
data may not be enough to fit a successful model. That said, we haven't looked
at the mini-challenge data, yet.

```{r}
rpdr %>% gather(key = "performance", value = "count", n_lipsync:n_wins) %>%
  filter(season == 4) %>%
  ggplot(aes(x = reorder(contestant, n_appearances), y = count, fill = fct_relevel(performance, "n_lipsync", "n_in_bottom", "n_safe", "n_in_top", "n_wins"))) +
  geom_col(alpha = 0.7) + 
  coord_flip() + 
  labs(y = "Count", fill  = "Performance", x = "Contestants", title = "Season 4") +
  theme_bw() + scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18), expand = c(0,0))
```

```{r}
rpdr %>% gather(key = "performance", value = "count", n_lipsync:n_wins) %>%
  filter(season == 5) %>%
  ggplot(aes(x = reorder(contestant, n_appearances), y = count, fill = fct_relevel(performance, "n_lipsync", "n_in_bottom", "n_safe", "n_in_top", "n_wins"))) +
  geom_col(alpha = 0.7) + 
  coord_flip() + 
  labs(y = "Count", fill  = "Performance", x = "Contestants", title = "Season 5") +
  theme_bw() + scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18), expand = c(0,0))
```

```{r}
rpdr %>% gather(key = "performance", value = "count", n_lipsync:n_wins) %>%
  filter(season == 6) %>%
  ggplot(aes(x = reorder(contestant, n_appearances), y = count, fill = fct_relevel(performance, "n_lipsync", "n_in_bottom", "n_safe", "n_in_top", "n_wins"))) +
  geom_col(alpha = 0.7) + 
  coord_flip() + 
  labs(y = "Count", fill  = "Performance", x = "Contestants", title = "Season 6") +
  theme_bw() + scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18), expand = c(0,0))
```

```{r}
rpdr %>% gather(key = "performance", value = "count", n_lipsync:n_wins) %>%
  filter(season == 7) %>%
  ggplot(aes(x = reorder(contestant, n_appearances), y = count, fill = fct_relevel(performance, "n_lipsync", "n_in_bottom", "n_safe", "n_in_top", "n_wins"))) +
  geom_col(alpha = 0.7) + 
  coord_flip() + 
  labs(y = "Count", fill  = "Performance", x = "Contestants", title = "Season 7") +
  theme_bw() + scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18), expand = c(0,0))
```

```{r}
rpdr %>% gather(key = "performance", value = "count", n_lipsync:n_wins) %>%
  filter(season == 8) %>%
  ggplot(aes(x = reorder(contestant, n_appearances), y = count, fill = fct_relevel(performance, "n_lipsync", "n_in_bottom", "n_safe", "n_in_top", "n_wins"))) +
  geom_col(alpha = 0.7) + 
  coord_flip() + 
  labs(y = "Count", fill  = "Performance", x = "Contestants", title = "Season 8") +
  theme_bw() + scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18), expand = c(0,0))
```

```{r}
rpdr %>% gather(key = "performance", value = "count", n_lipsync:n_wins) %>%
  filter(season == 9) %>%
  ggplot(aes(x = reorder(contestant, n_appearances), y = count, fill = fct_relevel(performance, "n_lipsync", "n_in_bottom", "n_safe", "n_in_top", "n_wins"))) +
  geom_col(alpha = 0.7) + 
  coord_flip() + 
  labs(y = "Count", fill  = "Performance", x = "Contestants", title = "Season 9") +
  theme_bw() + scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18), expand = c(0,0))
```

```{r}
rpdr %>% gather(key = "performance", value = "count", n_lipsync:n_wins) %>%
  filter(season == 10) %>%
  ggplot(aes(x = reorder(contestant, n_appearances), y = count, fill = fct_relevel(performance, "n_lipsync", "n_in_bottom", "n_safe", "n_in_top", "n_wins"))) +
  geom_col(alpha = 0.7) + 
  coord_flip() + 
  labs(y = "Count", fill  = "Performance", x = "Contestants", title = "Season 10") +
  theme_bw() + scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18), expand = c(0,0)) 
```

Looking across all of these graphs, a couple things stand out to me that are
worth exploring:

- 9/10 winners had or were tied for the most wins
- 5/10 winners never had to lipsync
- Of the winners who did not have the most wins, 3/6 never had to lipsync

This suggests to me that we could build a reasonably naive model based on wins
and lipsync data alone. So let's explore that quickly.

```{r}
rpdr.2 <- rpdr %>%
  group_by(season) %>%
  mutate(most_wins = ifelse(n_wins == max(n_wins), TRUE, FALSE),
  never_lipsynced = ifelse(n_lipsync == 0, TRUE, FALSE))
```

```{r}
model_nB <-naiveBayes(was_winner ~ most_wins + never_lipsynced, data = rpdr.2)
rpdr.2$prediction <- predict(model_nB, newdata = rpdr.2, type = "raw")

group_by(rpdr.2, season) %>%
  mutate(minval = min(prediction)) %>%
  filter(prediction == minval) %>%
  select(contestant)
```



```{r}
ggplot(rpdr, aes(x = n_lipsync, y = n_in_bottom, color = was_winner)) + 
  geom_point(position = position_jitter(w = 0.3, h = 0.3), alpha = 0.8) + 
  geom_hline(yintercept = 2, linetype = 2) +
  geom_vline(xintercept = 1, linetype = 2) +
  theme_bw()
```